---
title: "Shea et. al. Differential methylation plots (DMRs)"
author: "Deepak Tanwar"
date: "<b>Created on:</b> 2019-04-25 <br> <b>Updated on:</b> `r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: tango
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    keep_md: no
    number_sections: no
    fig_width: 8
    fig_height: 8
    fig_caption: true
    df_print: paged
    code_folding: show
  fontsize: 12pt
  geometry: margin=1in
  documentclass: article
# bibliography: references.bib
link-citations: yes
---


# Libraries

```{r, message=FALSE, warning=FALSE}
library(dmrseq)
library(ggplot2)
library(GenomicRanges)
library(plgINS)
library(imputeTS)
library(viridis)
library(pheatmap)
library(readxl)
library(data.table)
library(rtracklayer)
```


# Annotation track of mm10
```{r, message=FALSE, warning=FALSE}
annoTrack <- getAnnot("mm10")
```


# Import data
```{r, message=FALSE, warning=FALSE}
hf_ctrl <- readRDS("input/dmr_highFat_vs_ctrl.rds")
lp_ctrl <- readRDS("input/dmr_lowProtein_vs_ctrl.rds")
lp_hf <- readRDS("input/dmr_lowProtein_vs_highFat.rds")
```


# RRBS samples
```{r}
rrbs <- read.delim("./input/epigenomeDetails.mm9.2.RRBS.Promoter_centered_main_filtered.txt",
  header = T, sep = "\t", stringsAsFactors = F
)

rrbs.s <- rrbs[, c(1:6, grep(pattern = "methRatio", x = colnames(rrbs)))]

samples <- data.frame(read_xlsx("./input/Shea_2015_sigling_correlations.xlsx")[, 1])[-1, 1]
samples <- gsub("\\.", "_", samples)

final.rrbs <- data.frame(rrbs.s[, 1:6], rrbs.s[, grep(paste(samples, collapse = "|"), colnames(rrbs.s))])
final.rrbs <- na.replace(final.rrbs, 0)
final.rrbs <- GRanges(final.rrbs)

system("wget http://hgdownload.cse.ucsc.edu/goldenPath/mm9/liftOver/mm9ToMm10.over.chain.gz && gunzip mm9ToMm10.over.chain.gz")

chain <- import.chain("mm9ToMm10.over.chain")

rrbs.liftOver <- liftOver(x = final.rrbs, chain = chain)

final.rrbs <- rrbs.liftOver@unlistData
```


# Functions
## To create DMR plot
```{r, message=FALSE, warning=FALSE}
plot_regions <- function(dmr, regions = 1:10, groupsCol = "Group", anno = annoTrack) {
  for (i in regions) {
    name <- paste("Region", i)
    cat("\n \n")
    cat(paste("###", name))
    cat("\n \n")
    plotDMRs(
      BSseq = dmr$bs, regions = dmr$dmr[i, ], testCovariate = "Group",
      annoTrack = annoTrack, extend = 1000
    )
    cat("\n \n")
  }
}
```

## To create DMR table
```{r, message=FALSE, warning=FALSE}
make_DT <- function(dmr) {
  df <- data.frame(dmr$dmr[dmr$dmr$pval <= 0.05, ])[, 1:11]
  df[, 7:11] <- round(df[, 7:11], digits = 2)
  DT::datatable(
    df,
    rownames = F,
    filter = "top", extensions = c("Buttons", "ColReorder"), options = list(
      pageLength = 10,
      buttons = c("copy", "csv", "excel", "pdf", "print"),
      colReorder = list(realtime = FALSE),
      dom = "fltBip"
    )
  )
}
```

## To create histograms of `p-values`
```{r}
plot_hist <- function(dmr) {
  ggplot() + aes(dmr$dmr$pval) +
    geom_histogram(colour = "black", fill = "#3c8dbc", breaks = seq(0, 1, length.out = 20)) +
    ggtitle("Pvalue distribution") +
    xlab("P-value") + ylab("Frequency") +
    theme_classic() +
    theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 14)
    )
}
```

## To obtain M, Cov and Un matrices
```{r}
get_matrix <- function(bs, type = "all") {
  m <- getCoverage(bs, type = "M")
  c <- getCoverage(bs, type = "Cov")
  u <- c - m
  r <- data.frame(bs@rowRanges)
  beta <- data.frame(r, m / c)

  if (type == "M") {
    return(data.frame(r, m))
  } else if (type == "Cov") {
    return(data.frame(r, c))
  } else if (type == "Un") {
    return(data.frame(r, u))
  } else if (type == "beta") {
    return(beta)
  } else {
    return(list(M = data.frame(r, m), Cov = data.frame(r, c), Un = data.frame(r, u), Beta = beta))
  }
}
```

## To plot heatmap of regions
```{r}
plot_heatmap_regions <- function(dmrs, bs, cutoff = 0.05, cutoffCol = "pval",
                                 aggregateRegions = T, aggrFun = "mean", pCols,
                                 minOverlap = 1000, rrbs = NULL) {
  beta <- get_matrix(bs = bs, type = "beta")
  reg <- dmrs[data.frame(dmrs[, cutoffCol])[, cutoffCol] <= cutoff, ]

  overlap <- suppressWarnings(findOverlaps(subject = reg, query = GRanges(beta), type = "within"))
  res1 <- data.frame(reg[overlap@to, ], stringsAsFactors = F)
  res2 <- data.frame(GRanges(beta)[overlap@from], stringsAsFactors = F)
  res <- cbind(res1, res2)
  res <- res[!duplicated(res), c(1:11, 20:ncol(res))]
  res$ID <- paste(res[, 1], res[, 2], res[, 3], sep = "-")

  # overlap.M <- suppressWarnings(findOverlaps(subject = reg, query = GRanges(all$M), type = "within"))
  # res1.M <- data.frame(reg[overlap.M@to, ], stringsAsFactors = F)
  # res2.M <- data.frame(GRanges(all$M)[overlap.M@from], stringsAsFactors = F)
  # res.M <- cbind(res1.M, res2.M)
  # res.M <- res.M[!duplicated(res.M), c(1:11, 20:ncol(res.M))]
  # res.M$ID <- paste(res.M[, 1], res.M[, 2], res.M[, 3], sep = "-")
  #
  # overlap.Cov <- suppressWarnings(findOverlaps(subject = reg, query = GRanges(all$Cov), type = "within"))
  # res1.Cov <- data.frame(reg[overlap.Cov@to, ], stringsAsFactors = F)
  # res2.Cov <- data.frame(GRanges(all$Cov)[overlap.Cov@from], stringsAsFactors = F)
  # res.Cov <- cbind(res1.Cov, res2.Cov)
  # res.Cov <- res.Cov[!duplicated(res.Cov), c(1:11, 20:ncol(res.Cov))]
  # res.Cov$ID <- paste(res.Cov[, 1], res.Cov[, 2], res.Cov[, 3], sep = "-")

  beta <- res

  # if (aggregateRegions) {
  #   cov <- plag(x = res.Cov[, 12:(ncol(res.Cov) - 1)], by = res.Cov$ID, agFun = median)
  #   m <- plag(x = res.M[, 12:(ncol(res.M) - 1)], by = res.M$ID, agFun = median)
  #   b <- m / cov
  #   r <- res.Cov[, 1:11][!duplicated(res.Cov[, 1:11]), ]
  #   rownames(r) <- paste(r[, 1], r[, 2], r[, 3], sep = "-")
  #   beta <- data.frame(r, b)
  # } else {
  #   cov <- res.Cov[, 12:(ncol(res.Cov) - 1)]
  #   m <- res.M[, 12:(ncol(res.M) - 1)]
  #   b <- m / cov
  #   r <- res.Cov[, 1:11]
  #   beta <- data.frame(r, b)
  # }

  beta <- na.replace(beta, 0)
  beta <- beta[order(beta$pval), ]

  beta.a <- NULL

  if (aggregateRegions) {
    b <- plag(x = beta[, 12:(ncol(beta) - 1)], by = beta$ID, agFun = aggrFun)
    r <- beta[, 1:11][!duplicated(beta[, 1:11]), ]
    rownames(r) <- paste(r[, 1], r[, 2], r[, 3], sep = "-")
    beta.a <- data.frame(r, b)
  } else {
    beta.a <- beta
  }

  beta <- beta.a

  anno.row <- NULL
  anno.col <- NULL
  mat.plot <- NULL
  main <- NULL
  showRownames <- NULL

  if (dim(data.frame(rrbs))[1] > 0) {
    overlap <- suppressWarnings(findOverlaps(subject = GRanges(beta), query = rrbs, minoverlap = minOverlap))
    r1 <- data.frame(GRanges(beta)[overlap@to, ], stringsAsFactors = F)
    r2 <- data.frame(rrbs[overlap@from], stringsAsFactors = F)
    r <- cbind(r1, r2)

    if (dim(data.frame(r))[1] > 0) {
      r$ID <- paste0(as.character(r$seqnames), "-", r$start, "-", r$end, "(", r$geneNames, ")")
      r <- r[!duplicated(r$ID), ]
      rownames(r) <- r$ID
      r <- r[, -ncol(r)]

      anno.row <- data.frame(r[, c(4, 8:11)])
      anno.col <- data.frame(pData(bs)[, pCols], stringsAsFactors = F)

      s <- colnames(data.frame(rrbs))[grep("methRatio", colnames(data.frame(rrbs)))]
      df <- data.frame(ID = s, Age = "", Group = "RRBS")
      rownames(df) <- df$ID
      anno.col <- rbind(anno.col, df[, -1])

      g1 <- unique(anno.col$Group)[1]
      g2 <- unique(anno.col$Group)[2]
      g3 <- unique(anno.col$Group)[3]

      mat <- r
      mat1 <- mat[, rownames(anno.col)[anno.col$Group == g1]]
      mat2 <- mat[, rownames(anno.col)[anno.col$Group == g2]]
      mat3 <- mat[, rownames(anno.col)[anno.col$Group == g3]]
      mat.plot <- data.frame(mat1, mat2, mat3)

      main <- "DMRs beta values (aggregated) + RRBS"
      showRownames <- TRUE
    } else {
      anno.row <- beta[, c(4, 8:11)]
      anno.col <- data.frame(pData(bs)[, pCols], stringsAsFactors = F)

      g1 <- unique(anno.col$Group)[1]
      g2 <- unique(anno.col$Group)[2]

      mat <- beta[, 12:ncol(beta)]
      mat1 <- mat[, rownames(anno.col)[anno.col$Group == g1]]
      mat2 <- mat[, rownames(anno.col)[anno.col$Group == g2]]
      mat.plot <- data.frame(mat1, mat2)

      if (aggregateRegions) {
        main <- "DMRs beta values (aggregated)"
      } else {
        main <- "DMRs beta values (not aggregated)"
      }

      showRownames <- FALSE
    }
  } else {
    anno.row <- beta[, c(4, 8:11)]
    anno.col <- data.frame(pData(bs)[, pCols], stringsAsFactors = F)

    g1 <- unique(anno.col$Group)[1]
    g2 <- unique(anno.col$Group)[2]

    mat <- beta[, 12:ncol(beta)]
    mat1 <- mat[, rownames(anno.col)[anno.col$Group == g1]]
    mat2 <- mat[, rownames(anno.col)[anno.col$Group == g2]]
    mat.plot <- data.frame(mat1, mat2)

    if (aggregateRegions) {
      main <- "DMRs beta values (aggregated)"
    } else {
      main <- "DMRs beta values (not aggregated)"
    }

    showRownames <- FALSE
  }

  pheatmap(
    mat = mat.plot, color = viridis(n = 100), cluster_cols = F,
    show_rownames = showRownames, show_colnames = F,
    cluster_rows = F, annotation_row = anno.row,
    annotation_col = anno.col, border_color = NA,
    main = main
  )
}
```


# Histograms of `p-values` {.tabset .tabset-pills}

## HighFat vs CTRL
```{r, message=FALSE, warning=FALSE, fig.align='center', fig.height=5, fig.width=6}
plot_hist(dmr = hf_ctrl)
```

## LowProtein vs CTRL
```{r, message=FALSE, warning=FALSE, fig.align='center', fig.height=5, fig.width=6}
plot_hist(dmr = lp_ctrl)
```

## LowProtein vs HighFat
```{r, message=FALSE, warning=FALSE, fig.align='center', fig.height=5, fig.width=6}
plot_hist(dmr = lp_hf)
```



# DMR tables {.tabset .tabset-pills}

## HighFat vs CTRL
```{r, message=FALSE, warning=FALSE}
make_DT(hf_ctrl)
```

## LowProtein vs CTRL
```{r, message=FALSE, warning=FALSE}
make_DT(lp_ctrl)
```

## LowProtein vs HighFat
```{r, message=FALSE, warning=FALSE}
make_DT(lp_hf)
```


# DMR plots {.tabset .tabset-pills}

## HighFat vs CTRL {.tabset .tabset-dropdown}
```{r, results='asis', message=FALSE, warning=FALSE}
plot_regions(dmr = hf_ctrl, regions = 1:5, groupsCol = "Group", anno = annoTrack)
```

## LowProtein vs CTRL {.tabset .tabset-dropdown}
```{r, results='asis', message=FALSE, warning=FALSE}
plot_regions(dmr = lp_ctrl, regions = 1:5, groupsCol = "Group", anno = annoTrack)
```

## LowProtein vs HighFat {.tabset .tabset-dropdown}
```{r, results='asis', message=FALSE, warning=FALSE}
plot_regions(dmr = lp_hf, regions = 1:5, groupsCol = "Group", anno = annoTrack)
```


# Heatmap of `Beta` values

## HighFat vs CTRL {.tabset .tabset-dropdown}
### Aggregated
```{r, results='asis', message=FALSE, warning=FALSE, fig.align='center', fig.height=8, fig.width=8}
plot_heatmap_regions(
  dmrs = hf_ctrl$dmr, bs = hf_ctrl$bs,
  cutoff = 0.05, cutoffCol = "pval",
  aggregateRegions = T, aggrFun = "mean", pCols = 1:2,
  rrbs = NULL, minOverlap = 1000
)
```

### Aggregated + RRBS
```{r, results='asis', message=FALSE, warning=FALSE, fig.align='center', fig.height=8, fig.width=8}
plot_heatmap_regions(
  dmrs = hf_ctrl$dmr, bs = hf_ctrl$bs,
  cutoff = 0.05, cutoffCol = "pval",
  aggregateRegions = T, aggrFun = "mean", pCols = 1:2,
  rrbs = final.rrbs, minOverlap = 1000
)
```

### Not aggregated
```{r, results='asis', message=FALSE, warning=FALSE, fig.align='center', fig.height=8, fig.width=8}
plot_heatmap_regions(
  dmrs = hf_ctrl$dmr, bs = hf_ctrl$bs,
  cutoff = 0.05, cutoffCol = "pval",
  aggregateRegions = F, aggrFun = "mean", pCols = 1:2,
  rrbs = NULL, minOverlap = 1000
)
```


## LowProtein vs CTRL {.tabset .tabset-dropdown}
### Aggregated
```{r, results='asis', message=FALSE, warning=FALSE, fig.align='center', fig.height=8, fig.width=8}
plot_heatmap_regions(
  dmrs = lp_ctrl$dmr, bs = lp_ctrl$bs,
  cutoff = 0.05, cutoffCol = "pval",
  aggregateRegions = T, aggrFun = "mean", pCols = 1:2,
  rrbs = NULL, minOverlap = 1000
)
```

### Aggregated + RRBS
```{r, results='asis', message=FALSE, warning=FALSE, fig.align='center', fig.height=8, fig.width=8}
plot_heatmap_regions(
  dmrs = lp_ctrl$dmr, bs = lp_ctrl$bs,
  cutoff = 0.05, cutoffCol = "pval",
  aggregateRegions = T, aggrFun = "mean", pCols = 1:2,
  rrbs = final.rrbs, minOverlap = 1000
)
```

### Not aggregated
```{r, results='asis', message=FALSE, warning=FALSE, fig.align='center', fig.height=8, fig.width=8}
plot_heatmap_regions(
  dmrs = lp_ctrl$dmr, bs = lp_ctrl$bs,
  cutoff = 0.05, cutoffCol = "pval",
  aggregateRegions = F, aggrFun = "mean", pCols = 1:2,
  rrbs = NULL, minOverlap = 1000
)
```


## LowProtein vs HighFat {.tabset .tabset-dropdown}
### Aggregated
```{r, results='asis', message=FALSE, warning=FALSE, fig.align='center', fig.height=8, fig.width=8}
plot_heatmap_regions(
  dmrs = lp_hf$dmr, bs = lp_hf$bs,
  cutoff = 0.05, cutoffCol = "pval",
  aggregateRegions = T, aggrFun = "mean", pCols = 1:2,
  rrbs = NULL, minOverlap = 1000
)
```

### Aggregated + RRBS
```{r, results='asis', message=FALSE, warning=FALSE, fig.align='center', fig.height=8, fig.width=8}
plot_heatmap_regions(
  dmrs = lp_hf$dmr, bs = lp_hf$bs,
  cutoff = 0.05, cutoffCol = "pval",
  aggregateRegions = T, aggrFun = "mean", pCols = 1:2,
  rrbs = final.rrbs, minOverlap = 1000
)
```

### Not aggregated
```{r, results='asis', message=FALSE, warning=FALSE, fig.align='center', fig.height=8, fig.width=8}
plot_heatmap_regions(
  dmrs = lp_hf$dmr, bs = lp_hf$bs,
  cutoff = 0.05, cutoffCol = "pval",
  aggregateRegions = F, aggrFun = "mean", pCols = 1:2,
  rrbs = NULL, minOverlap = 1000
)
```


# SessionInfo
```{r}
devtools::session_info()
```
