---
title: "Differential methylation analysis (DMRs)"
author: "Deepak Tanwar"
date: "<b>Created on:</b> 2019-03-14 <br> <b>Updated on:</b> `r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: tango
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    keep_md: no
    number_sections: no
    fig_width: 8
    fig_height: 8
    fig_caption: true
    df_print: paged
    code_folding: show
    # toc: true
    # toc_depth: 5
  fontsize: 12pt
  geometry: margin=1in
  documentclass: article
# bibliography: references.bib
link-citations: yes
---

# Packages required

```{r, warning=F, message=F}
library(dmrseq)
library(DT)
library(readr)
library(forcats)
```

# Method description
In dmrseq, the methylation data are smoothed and then analyzes using generalized least squares model. Biological variability and correlation between neighbouring loci is taken into account.


# Annotation
```{r}
anno <- read.csv("./input/annotationRnBeads.csv", stringsAsFactors = F, row.names = 1)
anno$sampleID <- gsub(pattern = "-", replacement = "_", x = anno$sampleID)
rownames(anno) <- anno$sampleID
```

# Input files
Read in input files and create a `BSseq` object.

# Data
```{r}
bismarkBSseq <- read.bismark(
  files = paste0("./input/", anno$filename_bed),
  colData = anno,
  rmZeroCov = TRUE,
  strandCollapse = FALSE,
  verbose = TRUE,
  BPPARAM = BiocParallel::MulticoreParam(workers = 16)
)
```

# DMRs analysis

## Function
```{r}
perform_dmr <- function(bs, group = "Group", coVariate = NULL,
                        filterCol = NULL, filterColVals = NULL,
                        filt.loci = 0.4, nParallel = 2, revGroup = F) {
  bs.f <- NULL
  if (!is.null(filterCol)) {
    sample.idx <- which(pData(bs)[, filterCol] %in% c(filterColVals))
    bs.e <- bs[, sample.idx]
    bs.f <- bs.e[,which(!is.na(pData(bs.e)[, group]))]
  } else {
    bs.e <- bs
    bs.f <- bs.e[,which(!is.na(pData(bs.e)[, group]))]
  }

  g <- sort(unique(pData(bs.f)[,group]))
  
  loci.p <- floor((ncol(bs.f) * filt.loci) / 2)
  loci.idx <- which(rowSums(getCoverage(bs.f[, pData(bs.f)[, group] == g[1]], type = "Cov") > 0) >= loci.p &
    rowSums(getCoverage(bs.f[, pData(bs.f)[, group] == g[2]], type = "Cov") > 0) >= loci.p)

  bs.g <- bs.f[loci.idx, ]
  
  if(revGroup) pData(bs.g)[, group] <- fct_rev(pData(bs.g)[, group])

  dmr <- dmrseq(
    bs = bs.g,
    testCovariate = group,
    BPPARAM = BiocParallel::MulticoreParam(workers = nParallel),
    minNumRegion = 10,
    adjustCovariate = coVariate,
    cutoff = 0.1
  )
  r <- list(bs = bs.g, dmr = dmr)
  return(r)
}
```


## High-fat vs CTRL
```{r, message=FALSE, warning=FALSE}
dmr.hf.c <- perform_dmr(bs = bismarkBSseq, group = "Group1")
saveRDS(object = dmr.hf.c, file = "./output/dmr_highFat_vs_ctrl.rds")
```

<!-- ## High-fat vs CTRL [Covariate adjustment for age] -->
<!-- ```{r, message=FALSE, warning=FALSE} -->
<!-- dmr.hf.c.A <- perform_dmr(bs = bismarkBSseq, group = "Group1", coVariate = "Age") -->
<!-- saveRDS(object = dmr.hf.c.A, file = "./output/dmr_highFat_vs_ctrl_covar_age.rds") -->
<!-- ``` -->


## Low-protein vs CTRL
```{r, message=FALSE, warning=FALSE}
dmr.lp.c <- perform_dmr(bs = bismarkBSseq, group = "Group2")
saveRDS(object = dmr.lp.c, file = "./output/dmr_lowProtein_vs_ctrl.rds")
```

<!-- ## Low-protein vs CTRL [Covariate adjustment for age] -->
<!-- ```{r, message=FALSE, warning=FALSE} -->
<!-- dmr.lp.c.A <- perform_dmr(bs = bismarkBSseq, group = "Group2", coVariate = "Age") -->
<!-- saveRDS(object = dmr.lp.c.A, file = "./output/dmr_lowProtein_vs_ctrl_covar_age.rds") -->
<!-- ``` -->


## Low-protein vs High-fat
```{r, message=FALSE, warning=FALSE}
dmr.lp.hf <- perform_dmr(bs = bismarkBSseq, group = "Group3", revGroup = T)
saveRDS(object = dmr.lp.hf, file = "./output/dmr_lowProtein_vs_highFat.rds")
```

<!-- ## Low-protein vs High-fat [Covariate adjustment for age] -->
<!-- ```{r, message=FALSE, warning=FALSE} -->
<!-- dmr.lp.hf.A <- perform_dmr(bs = bismarkBSseq, group = "Group3", coVariate = "Age", revGroup = T) -->
<!-- saveRDS(object = dmr.lp.hf.A, file = "./output/dmr_lowProtein_vs_highFat_covar_age.rds") -->
<!-- ``` -->


<!-- # SaveImage -->
<!-- ```{r} -->
<!-- save.image(file = "output/20190315-dmrseq_analysis.RData", compress = T) -->
<!-- ``` -->

<!-- # DMR Table -->
<!-- Identified DMRs with q-value not higher than 0.05. -->
<!-- ```{r} -->
<!-- datatable( -->
<!--   dmr.nc, -->
<!--   rownames = F, -->
<!--   filter = "top", extensions = c("Buttons", "ColReorder"), options = list( -->
<!--     pageLength = 10, -->
<!--     buttons = c("copy", "csv", "excel", "pdf", "print"), -->
<!--     colReorder = list(realtime = FALSE), -->
<!--     dom = "fltBip" -->
<!--   ) -->
<!-- ) -->
<!-- ``` -->


# SessionInfo
```{r}
devtools::session_info()
```


```{r test, echo=F, eval=F}
dmr <- readRDS("output/dmr_lowProtein_vs_highFat.rds")

annoTrack <- getAnnot("mm10")

plotDMRs(dmr$bs, regions=dmr$dmr[1:10,], testCovariate="Group3",annoTrack=annoTrack)
```